First steps towards a formalization of Forcing

### 2018-06-23 LSFA/intro.tex
linea de más
edicion global

\subsection{Models of $\ZFC$}
By G\"odel's Second Incompleteness Theorem, we can't show prove that
there exists a model of $\ZFC$. More formally, if we assume that
mathematical proofs can be encoded as theorems of $\ZFC$ and that
the latter do not lead to contradictions (i.e., $\ZFC$ is
\emph{consistent}), then we can't prove that there exists a set $M$
and a binary relation $E$ such that $\lb M,E\rb$ satisfies the $\ZFC$
axioms.

A relative consistency proof for an axiom $A$ is then obtained by assuming that such a
$\lb M,E\rb$ exists, and constructing another $\lb M',E'\rb$
satisfying $\ZFC + A$. We single out a very special kind of models:
%
\begin{definition}\label{def:transitive-model}
  \begin{enumerate}
  \item A set $M$ (of sets) is \emph{transitive} if for all $x\in M$ and 
    $y\in x$, we have $y\in M$ (i.e., every element of $M$ is a subset
    of $M$).
  \item $\lb M,E\rb$ is a \emph{transitive model} if $M$ is transitive
    set and  $E$ is the membership relation $\in$ restricted to
    $M$. It is \emph{countable} if $M$ is coordinable with a subset of
    $\N$; we then say that $M$ is a \emph{ctm}. 
  \end{enumerate}
\end{definition}
%
As in the last sentence, one usually uses the underlying set to refer to a transitive
model, since the relation is fixed.

@@

Assume that $M$ is a ctm of $\ZFC$. In a portrayal convenient for this
introduction, G\"odel showed that $M$ contains a minimal submodel
$L^M$ of the same ``height'' (i.e.\ having the same ordinals)
that satisfies $\ZFC+\CH$. The sets in $L^M$ are called
\emph{constructible} and are in a sense ``definable.''

@@

\fbox{The next should be spread accross the other sections}

The easiest way to define genericity is by using preorders.
Given a preorder with top $\lb\PP,\leq,\1\rb$ in $M$, define
\emph{filter} $G$ ... , \emph{dense} ...  $G$ is an $M$-\emph{generic}
if it intersects every dense subset of $\PP$ that lies in $M$.

But since $M$ is countable,

The  Rasiowa-Sikorski theorem  states that that for every countable 
family of dense subsets there is a filter that intersects them
all. Thus, there are generic filters $G$ for countable transitive models.  In
general, no such  $G$ belongs to $M$ and therefore the extension
$M[G]$ is proper.

>>>>

\subsection{Models of $\ZFC$}
By G\"odel's Second Incompleteness Theorem, we can't  prove that
there exists a model of $\ZFC$. More formally, if we assume that
mathematical proofs can be encoded as theorems of $\ZFC$ and that
the latter do not lead to contradictions (i.e., $\ZFC$ is
\emph{consistent}), then we can't prove that there exists a set $M$
and a binary relation $E$ such that $\lb M,E\rb$ satisfies the $\ZFC$
axioms.

A relative consistency proof for an axiom $A$ is then obtained by assuming that such a
$\lb M,E\rb$ exists, and constructing another $\lb M',E'\rb$
satisfying $\ZFC + A$. We single out a very special kind of models:
%
\begin{definition}\label{def:transitive-model}
  \begin{enumerate}
  \item A set $M$ (of sets) is \emph{transitive} if for all $x\in M$ and 
    $y\in x$, we have $y\in M$ (i.e., every element of $M$ is a subset
    of $M$).
  \item $\lb M,E\rb$ is a \emph{transitive model} if $M$ is transitive
     and  $E$ is the membership relation $\in$ restricted to
    $M$. It is \emph{countable} if $M$ is coordinable with a subset of
    $\N$; we then say that the model  $M$ is a \emph{ctm}. 
  \end{enumerate}
\end{definition}
%
As in the last sentence, one usually uses the underlying set to refer to a transitive
model, since the relation is fixed.

@@

Assume that $M$ is a ctm of $\ZFC$. In a portrayal convenient for this
introduction, G\"odel showed that $M$ contains a minimal submodel
$L^M$ of the same ``height'' (i.e.\ having the same ordinals)
that satisfies $\ZFC+\CH$. The sets in $L^M$ are called
\emph{constructible} and are in a sense ``definable.'' In fact, there
is a first-order formula $L$ such that $L^M = \{x\in M : M\models
L(x)\}$.

@@

The easiest way to define genericity is by using a preorder with top
$\lb\PP,\leq,\1\rb$ in $M$.   
In Section~\ref{sec:forcing-notions} we formalize the definitions of
\emph{dense} subset and  \emph{filter} of  $\PP$, and we say that  $G$
is an $M$-generic filter
if it intersects every dense subset of $\PP$ that lies in $M$.

The  Rasiowa-Sikorski lemma  states that that for every countable 
family of dense subsets of a preorder there is a filter that intersects them
all. Thus, there are generic filters $G$ for countable transitive models.  In
general, no such  $G$ belongs to $M$ and therefore the extension
$M[G]$ is proper. We formalize the proof of the Rasiowa-Sikorski lemma
in Section~\ref{sec:rasiowa-sikorski-lemma}. A preliminary result on
a version of the Axiom of Choice is formalized in
Section~\ref{sec:sequence-version-dc}.

Elements of $M[G]$ are coded by the elements of $M$, and are obtained
from them by applying a function \isatt{val} that uses the generic filter
as a parameter. In order to show
that $M[G]$ does  extend $M$, we formalize the function \isatt{check}
that given an element $x\in M$ provides with a code (a ``name'') for
it.  This is explained in Section~\ref{sec:names}.

### 2018-06-23 LSFA/rs.tex
edicion global

\subsection{A sequence version of $\DC$}
In order to prove the Rasiowa-Sikorski lemma in
general, a version of the Axiom of Choice ($\AC$) must be proved. This
version is known as the \emph{Principle of Dependent choices}:
\begin{quote}
  ($\DC$) Let $\rr$ be a binary relation on $A$, and $a\in A$. If
  $\forall x\in A\,  \exists y\in A\, x\rr y$, then there exists
  $f:\om\to A$ such that $f(0)=a$ and $f(n)\rr f(n+1)$ for all
  $n\in\om$.
\end{quote}

@@

The function \verb|dc_witness| has, apart from $A$, $a$ and $R$, a
function $s$ as a parameter. If this function is a selector for
$\P(A)\sm \{\mty\}$, the function $f\defi$\verb|dc_witness|$(A,a,s,R)$
will satify $\DC$.

@@

\subsection{The Rasiowa-Sikorski lemma}
In order to state this Lemma, we gather the relevant hypotheses into a locale:

>>>>

\subsection{A sequence version of $\DC$}\label{sec:sequence-version-dc}
In order to prove the Rasiowa-Sikorski lemma in
general, a version of the Axiom of Choice ($\AC$) must be proved. This
version is known as the \emph{Principle of Dependent choices}:
\begin{quote}
  ($\DC$) Let $\rr$ be a binary relation on $A$, and $a\in A$. If
  $\forall x\in A\,  \exists y\in A\, x\rr y$, then there exists
  $f:\om\to A$ such that $f(0)=a$ and $f(n)\rr f(n+1)$ for all
  $n\in\om$.
\end{quote}

@@

The function \isatt{dc{\isacharunderscore}witness} has, apart from $A$, $a$ and $R$, a
function $s$ as a parameter. If this function is a selector for
$\P(A)\sm \{\mty\}$, the function $f\defi {}$\isatt{dc{\isacharunderscore}witness}$(A,a,s,R)$
will satify $\DC$. Notice that $s$ is a term of type \isatt{i} (a
function construed as a set of pairs) and an expression
\isatt{s{\isacharbackquote}b} is notation for  \isatt{apply(s,b)},
where
\isatt{apply\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}i\ {\isasymRightarrow}\ i\ {\isasymRightarrow}\ i{\isachardoublequoteclose}}
is the operation of function application. We will not make a
distinction in our prose and simply write \isatt{s}(\isatt{b}) in this case.

@@

\subsection{The Rasiowa-Sikorski lemma}\label{sec:rasiowa-sikorski-lemma}
In order to state this Lemma, we gather the relevant hypotheses into a locale:
